Model Comparisons

### 2016-01-12 models/Model_Comparison.Rmd
descending order for scale entropy measures
correlation plotting for langcog meeting on Tuesday 1.12.16

## Match items between studies (full and partial)
```{r}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "low1")))
```

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 16),
                                cor.e6=rep(NA, 16),
                                cor.e11=rep(NA, 16),
                                normalized=rep(NA, 16))
alphas = rep(NA, 16)
data.full.transfer = data.full[matched.items, ]
data.full.transfer$degree =
  ifelse(data.full.transfer$degree == "hi1", "hi", "low")
# Model 1 - Entailment only models
# --------------------------------
# 1.a)
# Entailment only
# Normalized = T
# alpha tuning = F
alphas[1] = 1
m1.a = data.partial %>%
  do(run.partial(., alpha = alphas[1], useprior=T, usenone=F, normalize=T))
# Store output and add e11
m1.a = cbind(m1.a, data.full.transfer$listener.p)
colnames(m1.a)[8] = "e11.listener.p"
performance.output[1, ] = c("m1.a", round(cor(m1.a$listener.p, m1.a$pred), 5),
                            round(cor(m1.a$e11.listener.p, m1.a$pred), 5), T)
# Store plot
m1.a.plot = qplot(stars, listener.p, col=degree, 
      data=m1.a,
      main=paste("m1.a\nalpha: ", alphas[1],
                 "\nNormalized = ", T,
                 "\nCorr = ", performance.output[1, 2]),
      ylab="Posterior p(rating | word)") + 
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

@@

# Store plot 2 - both e6 and e11 data
m4.d.jointD = m4.d.matched %>%
  gather(study, listener.pred, listener.p, e6.listener.p)
m4.d.jointD$study = ifelse(m4.d.jointD$study == "e6.listener.p", "e6", "e11")
m4.d.jointPlot = qplot(stars, listener.pred, col=degree, shape=study,
      data=m4.d.jointD,
      main=paste("m4.d\nalpha: ", alphas[16],
                 "\nNormalized = ", T,
                 ylab="Posterior p(rating | word)")) +
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
```

@@

@@

Across e6 and e11 pragmatic listener judgements we see model fit improve as we add more alternatives: The largest improvement occurs when we first include the empirically derived salient alternatives (correlations jump +0.23). Have a look at the `m2` vs `m3` correlation for models with pre-RSA normalization below (`c` models)
```{r}
# Tuned, normalized models r
barchart.d = performance.output[c(3, 7, 11, 15), ] %>%
  gather(study, cor, cor.e6, cor.e11)
barchart.d$cor = as.numeric(barchart.d$cor)
ggplot(barchart.d, aes(x=model, y=cor, fill=study)) + 
  geom_bar(stat="identity", position="dodge") +
  ylim(0, 1) + 
  ggtitle("Model fit for e6 and e11\npragmatic listener judgments") +
  geom_text(aes(x=model, y=as.numeric(cor) + 0.025, label=round(cor, 3), col=(study)))
```

@@

How can we interpret some of the more funky alternatives (i.e. `memorable_unforgettable`, `some_all`)? Does this scalar family 'want' more than one more alternative? That is, maybe the model is telling us that only one additional alternative is not enough...? Can we link this intuition back to entropy measures for the scales? 
```{r}
# entropy measures 
scales.entropy
```

@@

m2.a.plot
m2.b.plot
m2.c.plot
m2.d.plot

>>>>

## Match items between studies (full and partial)
```{r}
matched.items = which((data.full[, "scale"] != "some_all" &
        (data.full[, "degree"] == "hi2" | data.full[, "degree"] == "hi1")) |
       (data.full[, "scale"] == "some_all" &
          (data.full[, "degree"] == "hi1" | data.full[, "degree"] == "low1")))
# Run if looking at MLE alts
# --------------------------
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "low1")))
```

## Run comparisons
```{r run_comparisons, message=FALSE}
# Save performance output
performance.output = data.frame(model=rep(NA, 20),
                                cor.e6=rep(NA, 20),
                                cor.e11=rep(NA, 20),
                                normalized=rep(NA, 20))
alphas = rep(NA, 20)
data.full.transfer = data.full[matched.items, ]
data.full.transfer$degree =
  ifelse(data.full.transfer$degree == "hi1", "hi", "low")
# Model 1 - Entailment only models
# --------------------------------
# 1.a)
# Entailment only
# Normalized = T
# alpha tuning = F
alphas[1] = 1
m1.a = data.partial %>%
  do(run.partial(., alpha = alphas[1], useprior=T, usenone=F, normalize=T))
# Store output and add e11
m1.a = cbind(m1.a, data.full.transfer$listener.p)
colnames(m1.a)[8] = "e11.listener.p"
performance.output[1, ] = c("m1.a", round(cor(m1.a$listener.p, m1.a$pred), 5),
                            round(cor(m1.a$e11.listener.p, m1.a$pred), 5), T)
# Store plot
m1.a.plot = qplot(stars, listener.p, col=degree, 
      data=m1.a,
      main=paste("m1.a\nalpha: ", alphas[1],
                 "\nNormalized = ", T,
                 "\nCorr = ", performance.output[1, 2]),
      ylab="Posterior p(rating | word)") + 
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

@@

# Store plot 2 - both e6 and e11 data
m4.d.jointD = m4.d.matched %>%
  gather(study, listener.pred, listener.p, e6.listener.p)
m4.d.jointD$study = ifelse(m4.d.jointD$study == "e6.listener.p", "e6", "e11")
m4.d.jointPlot = qplot(stars, listener.pred, col=degree, shape=study,
      data=m4.d.jointD,
      main=paste("m4.d\nalpha: ", alphas[16],
                 "\nNormalized = ", T,
                 ylab="Posterior p(rating | word)")) +
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

# Model 5 - Empirical alternatives + Empircal extras (e12)
# -----------------------------------------

# Run if looking at e12 data
# --------------------------
matched.items.extras = which((data.full.extras[, "scale"] != "some_all" &
        (data.full.extras[, "degree"] == "hi2" | data.full.extras[, "degree"] == "hi1")) |
       (data.full.extras[, "scale"] == "some_all" &
          (data.full.extras[, "degree"] == "hi1" | data.full.extras[, "degree"] == "mid")))

speaker = read.csv("~/Desktop/Projects/scalar_implicature/models/model_data/L0_e12.csv")
data.full.extras = left_join(speaker, listener) %>%
  left_join(priors) %>%
  rowwise %>%
  select(scale, degree, stars, speaker.p, listener.p, prior.p) %>%
  mutate(listener.p = ifelse(is.na(listener.p), 0, listener.p)) %>%
  group_by(scale)

# 5.a)
# Emp alts + extras
# Normalized = T
# alpha tuning = F
alphas[17] = 1
m5.a = data.full.extras %>%
  do(run.full(., alpha = alphas[17], addMid=T, normalize=T))
# Match with e6 w/ checks
m5.a.matched = m5.a[matched.items.extras, ]
m5.a.matched$degree = ifelse(m5.a.matched$degree == "hi1", "hi", "low")
m5.a.matched = cbind(m5.a.matched, data.partial$listener.p)
all(m5.a.matched$scale == data.partial$scale & m5.a.matched$degree == data.partial$degree)
colnames(m5.a.matched)[length(colnames(m5.a.matched))] = "e6.listener.p"
# Store output
performance.output[17, ] = c("m5.a", round(cor(m5.a.matched$e6.listener.p, m5.a.matched$pred), 5), 
                             round(cor(m5.a.matched$listener.p, m5.a.matched$pred), 5), T)
# Store plot
m5.a.plot = qplot(stars, e6.listener.p, col=degree, 
      data=m5.a.matched,
      main=paste("m5.a\nalpha: ", alphas[17],
                 "\nNormalized = ", T,
                 "\nCorr = ", performance.output[17, 2]),
      ylab="Posterior p(rating | word)") + 
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

# 5.b)
# Emp alts + extras
# Normalized = F
# alpha tune = F
alphas[18] = 1
m5.b = data.full.extras %>%
  do(run.full(., alpha = alphas[18], addMid=T, normalize=F))
# Match with e6 w/ checks
m5.b.matched = m5.b[matched.items.extras, ]
m5.b.matched$degree = ifelse(m5.b.matched$degree == "hi1", "hi", "low")
m5.b.matched = cbind(m5.b.matched, data.partial$listener.p)
all(m5.b.matched$scale == data.partial$scale & m5.b.matched$degree == data.partial$degree)
colnames(m5.b.matched)[length(colnames(m5.b.matched))] = "e6.listener.p"
# Store output
performance.output[18, ] = c("m5.b", round(cor(m5.b.matched$e6.listener.p, m5.b.matched$pred), 5), 
                             round(cor(m5.b.matched$listener.p, m5.b.matched$pred), 5), F)
# Store plot
m5.b.plot = qplot(stars, e6.listener.p, col=degree, 
      data=m5.b.matched,
      main=paste("m5.b\nalpha: ", alphas[18],
                 "\nNormalized = ", F,
                 "\nCorr = ", performance.output[18, 2]),
      ylab="Posterior p(rating | word)") + 
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

# 5.c)
# Emp alts + extras
# Normalized = T
# alpha tune = T
# compare.data = Toggle this with data.partial or NULL
alphas[19] = tune.alpha(data.full.extras, type = "full", compare.data=data.partial, addMid=T, normalize=T)
m5.c = data.full.extras %>%
  do(run.full(., alpha = alphas[19], addMid=T, normalize=T))
# Match with e6 w/ checks
m5.c.matched = m5.c[matched.items.extras, ]
m5.c.matched$degree = ifelse(m5.c.matched$degree == "hi1", "hi", "low")
m5.c.matched = cbind(m5.c.matched, data.partial$listener.p)
all(m5.c.matched$scale == data.partial$scale & m5.c.matched$degree == data.partial$degree)
colnames(m5.c.matched)[length(colnames(m5.c.matched))] = "e6.listener.p"
# Store output (e11 and e6)
performance.output[19, ] = c("m5.c", round(cor(m5.c.matched$e6.listener.p, m5.c.matched$pred), 5),
                             round(cor(m5.c.matched$listener.p, m5.c.matched$pred), 5), T)

# Store plot 2 - both e6 and e11 data
m5.c.jointD = m5.c.matched %>%
  gather(study, listener.pred, listener.p, e6.listener.p)
m5.c.jointD$study = ifelse(m5.c.jointD$study == "e6.listener.p", "e6", "e11")
m5.c.jointPlot = qplot(stars, listener.pred, col=degree, shape=study,
      data=m5.c.jointD,
      main=paste("m5.c\nalpha: ", alphas[19],
                 "\nNormalized = ", T,
                 ylab="Posterior p(rating | word)")) +
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)

# 5.d)
# Emp alts + extras
# Normalized = F
# alpha tune = T
# compare.data = Toggle this with data.partial or NULL
alphas[20] = tune.alpha(data.full.extras, type = "full", compare.data=data.partial, addMid=T, normalize=F)
m5.d = data.full.extras %>%
  do(run.full(., alpha = alphas[20], addMid=T, normalize=F))
# Match with e6 w/ checks
m5.d.matched = m5.d[matched.items.extras, ]
m5.d.matched$degree = ifelse(m5.d.matched$degree == "hi1", "hi", "low")
m5.d.matched = cbind(m5.d.matched, data.partial$listener.p)
all(m5.d.matched$scale == data.partial$scale & m5.d.matched$degree == data.partial$degree)
colnames(m5.d.matched)[length(colnames(m5.d.matched))] = "e6.listener.p"
# Store output
performance.output[20, ] = c("m5.d", round(cor(m5.d.matched$e6.listener.p, m5.d.matched$pred), 5),
                             round(cor(m5.d.matched$listener.p, m5.d.matched$pred), 5), F)

# Store plot 2 - both e6 and e11 data
m5.d.jointD = m5.d.matched %>%
  gather(study, listener.pred, listener.p, e6.listener.p)
m5.d.jointD$study = ifelse(m5.d.jointD$study == "e6.listener.p", "e6", "e11")
m5.d.jointPlot = qplot(stars, listener.pred, col=degree, shape=study,
      data=m5.d.jointD,
      main=paste("m5.d\nalpha: ", alphas[20],
                 "\nNormalized = ", T,
                 ylab="Posterior p(rating | word)")) +
  ylim(0, 1) +
  facet_wrap(~scale) + 
  geom_line(aes(y = pred), lty = 4)
```

## Looking at empirically derived 'mids'
```{r}
normalize = function(w) { return (w / sum(w))}

scalars = c("good_excellent",
            "liked_loved",
            "palatable_delicious",
            "memorable_unforgettable",
            "some_all")

for (scale in scalars) {
  degrees = unique(data.full.extras$degree)
  for (degree in degrees) {
    data.full.extras[data.full.extras$scale == scale &
                       data.full.extras$degree == degree, ]$speaker.p = 
        normalize(data.full.extras[data.full.extras$scale == scale &
                                     data.full.extras$degree == degree, ]$speaker.p)
  }
}

# Differentiate tuned 'mid' scalar
plot.data = data.full.extras
plot.data$type = ifelse((plot.data$degree == "mid" & plot.data$scale != "some_all") |
                          (plot.data$degree == "low1" & plot.data$scale == "some_all"), "red", "blue")
qplot(x = stars, y = speaker.p, colour = type,
      geom = "line", stat = "identity", position = "dodge", 
      main = "Literal listener semantics\nEmpirical estimates and tuned extra 'mid'",
      ylab = "P_{L0}(Word | Rating) - Normalized",
      data = plot.data) + 
  facet_wrap(~scale + degree)
```

@@

Output for LangCog Lab meeting 1/12/16
```{r}
# Relevant studies into long data
plot.data <- subset(performance.output, model %in% c("m1.c", "m2.c", "m3.c"))
plot.data <- plot.data %>%
  gather(exp, cor, -c(normalized, model))
ggplot(plot.data, aes(x=model, y=cor, fill=exp)) +
  geom_bar(stat="identity", position="dodge") +
  ylim(0.0, 1.0)
```

@@

Across e6 and e11 pragmatic listener judgements we see model fit improve as we add more alternatives: The largest improvement occurs when we first include the empirically derived salient alternatives (correlations jump +0.23). Have a look at the `m2` vs `m3` correlation for models with pre-RSA normalization below (`c` models)
```{r}
# Tuned, normalized models r
barchart.d = performance.output[c(3, 7, 11), ] %>%
  gather(study, cor, cor.e6, cor.e11)
barchart.d$model <- as.factor(barchart.d$model)
levels(barchart.d$model) <- c("Entailment only",
                              "Entailment + \ngeneric none",
                              "Full model")
barchart.d$cor = as.numeric(barchart.d$cor)
cor.chart <- ggplot(barchart.d, aes(x=model, y=cor, fill=study)) + 
  geom_bar(stat="identity", position="dodge") +
  ylab("Correlation") +
  xlab("Model") +
  ylim(0, 1) + 
  ggtitle("Model fit correlation coefficients") +
  guides(colour=FALSE) +
  geom_text(aes(x=model, y=as.numeric(cor) + 0.025, label=round(cor, 3), col=(study)))

png(filename="practiceChart.png", 
    #type="cairo",
    units="in", 
    width=10, 
    height=8, 
    pointsize=12, 
    res=96)
cor.chart
dev.off()
```

@@

How can we interpret some of the more funky alternatives (i.e. `memorable_unforgettable`, `some_all`)? Does this scalar family 'want' more than one more alternative? That is, maybe the model is telling us that only one additional alternative is not enough...? Can we link this intuition back to entropy measures for the scales? 
```{r}
# entropy measures 
scales.entropy[order(scales.entropy$Entropy, decreasing=T),]
```

@@

m2.a.plot
m2.b.plot
m2.d.plot
m2.c.plot
